Interisland movements of woodland caribou in Newfoundland

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd and manuscript.md

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond_2016]. We present two
explanations for these swimming events, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007), and discuss why these swimming events
appear to be relatively commonplace in the population.

@@

## Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

The forage limitation and predator avoidance hypotheses are generally thought to
apply uniformly across populations, but individual caribou could evaluate the
costs and benefits of swimming to another island differently. Potentially those
three females, compared to the other 26, considered swimming less costly or
perceived competition or predation to be of greater concern. The individual
attributes that might influence how animals vary in their assessments of these
trade-offs are unknown, and potential predictors such as age or personality
traits are unknown for these caribou.

## Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris_1987]. While this is an ultimate
explanation to a series of proximate observations, it is possible for
density-dependent habitat selection to operate at such a fine-scale (Webber &
Vander Wal 2018). Evidence from the Sable Island horse (Equus ferus caballus)
system corroborates this hypothesis (van Beest et al. 2014). For horses,
selection for high quality habitat was strongest at low population densities,
but individuals tended to settle in lower quality habitats as population density
increased, following Ideal Free Distribution theory (van Beest et al. 2014).
Predator avoidance can also contribute to caribou swimming behaviour, but in
this situation it seems a less likely explanation given predation is primarily a
concern for young calves, which are the least likely to engage in this costly
behaviour.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible?-->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002; Jeffery
et al. 2007; Ricca et al. 2012) have noted the exceptional nature of caribou
swimming in the ocean, we suggest that this behaviour is likely more common than
previously thought for caribou living on oceanic islands. We posit forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for why caribou swimming in the ocean, with predator avoidance as a
potential proximate explanation. Although our inference is limited to
observations, our observations suggest that islands represent discrete foraging
patches for terrestrial animals that could influence fitness via increased
foraging opportunities and safety from predators. We also suggest that
individuals vary in their evaluation of costs and benefits of movement between
these habitat patches.

>>>>

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond_2016]. We present two
explanations for these swimming events, forage limitation (Miller 2002) and
predator avoidance [@Jeffrey_2007], and discuss why these swimming events
appear to be relatively commonplace in the population.

@@

## Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. 
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

The forage limitation and predator avoidance hypotheses are generally thought to
apply uniformly across populations, but individual caribou could evaluate the
costs and benefits of swimming to another island differently. It is possible
that some individuals, in our case ~10%, considered swimming less costly or that
they perceived competition or predation more acutely than their conspecifics.
Intrinsic life-history, physiological, or behavioural traits that might
influence how animals vary in their assessments of swimming trade-offs remains
unknown. Future work could assess the role of age, stress, or an individuals
tendency to move potential predictors of swimming behaviour.

## Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris_1987]. While this is an
ultimate explanation to a series of proximate observations, fine-scale
density-dependent habitat selection is theoretically possible
[@Webber_2017]. For horses (*Equus ferus caballus*), selection for high quality
habitat was strongest at low population densities, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory [@van_Beest_2013]. Predator avoidance could also
contribute to caribou swimming behaviour, but given the presumed low density of coyotes on Fogo Island and that predation primarily only affects young calves, we suspect predator is a possible, but less likely explanation.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit -->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean (Table 1). While past observations (e.g. @Miller_1995;
@Jeffrey_2007; @Ricca_2012] have noted the exceptional nature of caribou
swimming in the ocean, we suggest that this behaviour is likely more common than
previously thought for caribou living on oceanic islands. We posit forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for why caribou swimming in the ocean, with predator avoidance as a
potential proximate explanation. Although our inference is limited to
observations, our observations suggest that islands represent discrete foraging
patches for terrestrial animals that could influence fitness via increased
foraging opportunities and safety from predators. We also suggest that
individuals vary in their evaluation of costs and benefits of movement between
these habitat patches.
